---
title: "Correlation Homework"
author: "Kyleigh Johnson"
date: 11/14/24
format: html
editor: visual
---

*This is my completion of the correlation homework using Rice Rivers Center water quality data.*

## Look up the library GGall1y; it has a function named ggpairs(). Use that function to plot the atmospheric data from the Rice Rivers Center for the samples collected from 10 February to 15 February.

```{r}
source( "getRiceData.R" )
rice <- getRiceData()

library( GGally )

```

```{r}
rice |>
  filter( Month == "February",
          Day >= 10, 
          Day <= 15)  |>
  reframe( AirTempC, WindSpeed_kph, WindDir, RelHumidity, BP_HG )  -> df

ggpairs( df )

```

## For those atmospheric data, which pair of variables has the strongest correlation? What is the 95% confidence interval on that correlation coefficient?

Barometric Pressure and Air Temperature seem to have the strongest (negative) correlation with a r number of -0.712.

```{r}

cor_test_result <- cor.test( rice$AirTempC, rice$BP_HG )

cor_test_result
```

The 95% confidence interval is -0.5495821 and -0.5186443.

## Using the first 40 observations in air temperature and barometric pressure from the Rice Center data set, determine if they are individually distributed as normal random variables.

**Barometric Pressure**

```{r}
qqnorm( rice$BP_HG[1:40] )
qqline( rice$BP_HG[1:40], col="red")
```

```{r}
shapiro.test( rice$BP_HG[1:40] )
```

This data does not appear to be normal (we reject the null hypothesis).

**Air Temperature**

```{r}
qqnorm( rice$AirTempC[1:40] )
qqline( rice$AirTempC[1:40], col="red")
```

```{r}
shapiro.test( rice$AirTempC[1:40] )
```

This data appears to be normal (we fail to reject the null hypothesis).

## Given your findings in the last question, what kind of correlation statistic would be most appropriate for estimating the correlation between this subset of data?

Since one of the correlations is non-normal, I would think to use the Spearman's Rho coefficient.

```{r}
cor.test( rice$AirTempC, rice$BP_HG, 
                            method = "spearman" )
```

## Look at a qqnorm() plot of the barometric pressure data you used in the previous example. Is there something that "looks" odd with these data? Explain why those data are the way they are.

This plot appears to be staggered, giving it a stair-like appearance, rather than a more "normal" straight line. This is probably because of limitations of the instruments used and environmental data is unpredictable.

## Using a permutation approach, define the distribution of correlation values between the variables in #3, assuming that the NULL hypothesis is true. Plot these as a histogram and include the observed correlation.

```{r}
cor.test( rice$AirTempC[1:40], rice$BP_HG[1:40], 
                            method = "spearman" )
```

```{r}
df <- data.frame( Estimate = factor( c( "Original",
                                        rep("Permuted", 999))), 
                  rho =  c( cor.test( rice$AirTempC[1:40], rice$BP_HG[1:40] )$estimate,
                            rep(NA, 999)) )

summary( df )
```

```{r}
for( i in 2:1000) {
  yhat <- sample( rice$BP_HG,
                  size = length(rice$BP_HG[1:40]), 
                  replace = FALSE)
  model <- cor.test( rice$AirTempC[1:40], yhat )
  df$rho[i] <- model$estimate 
}
```

```{r}
ggplot( df ) + 
  geom_histogram( aes(rho, fill=Estimate ) )+
  labs( title = "Permutation Distribution",
        x = "Spearman",
        y = "Frequency")
```

The original value is smaller than most of the permuted values.
