---
title: "Factors"
format: html
editor: visual
author: Kyleigh Johnson
date: 10/24/24
---

This is an example of using the `forcats` library to manipulate categorical data.

## 1. Change the order of the weekdays to start with Monday.

First, I load in the proper libraries, my rice data function, and assign the `rice` variable.

```{r}
#| message: false
#| warning: false 

source( "getRiceData.R" )

rice <- getRiceData()

library( forcats )

library( knitr )

library( kableExtra )

library( lunar )

```

Next, I use the `forcats` library to reorder the data weekdays to begin with Monday.

```{r}
rice$Weekday <- factor( rice$Weekday )

rice <- rice |>
  mutate( Weekday = fct_relevel( Weekday,
                                 c( "Monday", "Tuesday", "Wednesday", "Thursday", 
                                   "Friday", "Saturday", "Sunday" )))
levels( rice$Weekday )

```

## 2. Add two additional factor columns: Weekday/Weekend based on the day of the week and create a column based on the *four primary lunar phases* for the data collected.

Here I added a factor column for the Weekday/Weekend.

```{r}
wkdy_wknd <- function( day ) {
  if (day %in% c( "Saturday", "Sunday" ) ) {
    return( "Weekend" )
  } else {
    return( "Weekday" )
  }
}

df <- rice |>
  mutate(Type_Of_Day = ifelse( Weekday %in% c("Saturday", "Sunday"), "Weekend", "Weekday" ) )

df |>
  head() |>
  kable() |>
  kable_minimal()
```



Next, I add in a column stating the lunar phase.

```{r}
df |>
  mutate( Lunar_Phase = lunar.phase( Date, shift = 0, name = 4)) -> df2


df2 |>
  head() |>
  kable() |>
  kable_minimal()

  
```

## 3. [Superstar Question] Create a column that indicates whether the tide is coming in (Incoming) or going out (Outgoing).



Last, I added a column to show whether the tide was incoming or outgoing.

```{r}
df2 |>
  mutate( diff_depth = Depth_m - lag( Depth_m ) ) |>
  mutate( Tide = if_else( diff_depth > 0, "Incoming", "Outgoing")) |>
  select( -diff_depth )-> df3


df3 |>
  head() |>
  kable() |>
  kable_minimal()
```

I then reordered the table so that the three added columns were not at the end to show for the purpose of this activity.

```{r}
df3 |>
  select( Date, Month, Day, Weekday, Type_Of_Day, Lunar_Phase, Tide, everything() ) |>
  head() |>
  kable() |>
  kable_minimal()

```
