---
title: "Factors"
format: html
editor: visual
author: Kyleigh Johnson
date: 10/24/24
---
This is an example of using the `forcats` library to manipulate categorical data.

## 1.Change the order of the weekdays to start with Monday.
First, I load in the proper libraries, my rice data function, and assign the `rice` variable.
```{r}
#| message: false
#| warning: false 

source( "GetRiceData.R" )

rice <- getRiceData()

library( forcats )

library( knitr )

library( kableExtra )

```
Next, I use the `forcats` library to reorder the data weekdays to begin with Monday.

```{r}
weekdays <- c("Monday","Tuesday","Wednesday",
              "Thursday","Friday","Saturday", 
              "Sunday")
class( weekdays )

data <- sample( weekdays, size=40, replace=TRUE)
data

days <- factor( data )
is.factor( days )

data <- factor( days, ordered=TRUE, levels = weekdays )
data

sort( data )

```
## 2.Add two additional factor columns: Weekday/Weekend based on the day of the week and create a column based on the *four primary lunar phases* for the data collected.

Here I added a factor column for the Weekday/Weekend.

```{r}
wkdy_wknd <- function( day ) { if
  ( day %in% c( "Saturday", "Sunday" ) ) {
    return( "Weekend" )
  } else {
    return( "Weekday" )
  }
}

wkdy_O_wknd <- factor( sapply ( days, wkdy_wknd ),
                       levels = c( "Weekday", "Weekend" ) )
data.frame(Days = days, Weekday_Weekend = wkdy_O_wknd) -> df

kable_minimal()
```

