---
title: "Graphics_HW"
format: html
editor: visual
author: Kyleigh Johnson
date: 10/15/24
---
This is a quick demonstration of data manipulation and the creation of different graphs and charts using basic R plots and the ggplot2 package.

# Load in Data

```{r}
source( "GetRiceData.R" )

rice <- getRiceData()
```

# Create a Bar Graph for Rain on Weekdays in January Using Basic Plots

```{r}

rice |>
  filter( Month == "January") |>
  group_by( Weekday ) |>
  summarize( Rain_cm = sum( Rain_cm ) ) -> df

barplot( df$Rain_cm,
         names.arg = df$Weekday,
         ylab = "Rain (cm)",
         xlab = "Weekday",
         cex.names = 0.7,
         main = "Rain on Weekdays in January")

```

# Create a Bar Graph for Rain on Weekdays in January Using ggplot2

```{r}
library( ggplot2 )

rice |>
  filter( Month == "January") |>
  group_by( Weekday ) |>
  summarize( Rain_cm = sum( Rain_cm ) )

ggplot( rice, aes( x = Weekday, Rain_cm) ) + 
  geom_col() +
  labs( title = "Rain on Weekdays in January") +
  theme(plot.title = element_text(hjust = 0.5))

```

# Create a Bar Graph for Temp on Weekdays in February Using Basic Plots

```{r}

rice |>
  filter( Month == "February") |>
  group_by( Weekday ) |>
  summarize( temps = mean( AirTempC ) ) -> df

barplot( df$temps,
      names.arg = df$Weekday,
      ylab = "Air Temp (C)",
      xlab = "Weekday",
      cex.names = 0.7,
      main = "Average Air Temp on Weekdays in February")

```

# Create a Bar Graph for Temp on Weekdays in February Using ggplot2

```{r}
rice |>
  filter( Month == "February") |>
  group_by( Weekday ) |>
  summarize( temps = mean( AirTempC ) )

ggplot( df, aes( x = Weekday, y = temps) ) + 
  geom_col() +
  labs( title = "Average Air Temp on Weekdays in February") +
  theme(plot.title = element_text(hjust = 0.5))

```

# Create a Line Graph for Water Depth on Valentine's Day Using Basic Plots

```{r}
rice |>
  filter( Month == "February",
          Day == "14") |>
  group_by( Date ) |>
  summarize( depth = mean( Depth_m ),
             date = Date )  -> df

plot( df$date, df$depth,
      type = "l",
      ylab = "Depth (m)",
      xlab = "Time",
      main = "Water Depth on Valentine's Day")

```

# Create a Line Graph for Water Depth on Valentine's Day Using ggplot2

```{r}
rice |>
  filter( Month == "February",
          Day == "14") |>
  group_by( Date ) |>
  summarize( depth = mean( Depth_m ) )  -> df

ggplot( df, aes( x = Date, y = depth) ) + 
  geom_line(color = "blue", linewidth = 1.2) +         
  labs(title = "Water Depth on Valentine's Day",    
       x = "Time",                                
       y = "Depth (m)") +                        
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

# Create a Scatter Plot for Air Temp vs Rainfall Using Basic Plots

```{r}
rice |>
  group_by( Date ) |>
  summarize( temp = median( AirTempC, na.rm = TRUE ),
             rain = sum( Rain_cm, na.rm = TRUE )) -> df

df |>
  group_by(Month = lubridate::month(Date, label = TRUE)) |>
  summarize(mean_temp = mean(temp, na.rm = TRUE),
            mean_rain = mean(rain, na.rm = TRUE)) -> centroids

plot(df$temp, df$rain, 
     main = "Median Temp vs Total Rainfall",
     xlab = "Median Air Temp (C)",
     ylab = "Total Rainfall (cm)",
     pch = 19,
     frame = FALSE,
     ylim = c( 0, max( df$rain ) * 1.5))

points(centroids$mean_temp, centroids$mean_rain, 
       col = "red",                 
       pch = 17,                   
       cex = 1)
```

# Create a Scatter Plot for Air Temp vs Rainfall Using ggplot2

```{r}
rice |>
  group_by( Date ) |>
  summarize( temp = median( AirTempC, na.rm = TRUE ),
             rain = sum( Rain_cm, na.rm = TRUE )) -> df

df |>
  group_by(Month = lubridate::month(Date, label = TRUE)) |>
  summarize(mean_temp = mean(temp, na.rm = TRUE),
            mean_rain = mean(rain, na.rm = TRUE)) -> centroids

ggplot( df, aes( x = temp, y = rain)) +
  geom_point() +
  geom_point( data = centroids, aes(x = mean_temp, y = mean_rain),
              color = "red", size = 2, shape = 16) +
  labs( title = "Median Temp vs Total Rainfall",
        x = "Median Air Temp (C)",
        y = "Total Rainfall (cm)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))


```

# Create a Pie Chart for Wind Direction (Week of Jan 8) Using Basic Plots

```{r}

rice |>
   filter(month( Date ) == 1 & day( Date ) >= 7 & day( Date ) <= 13 ) |>
  group_by(Date) |>
  summarize(wind_direction = mean(WindDir, na.rm = TRUE)) -> df
  
breaks <- seq(0, 360, by = 45)  
df$wind_direction_category <- cut(df$wind_direction, breaks = breaks, right = FALSE)

category_counts <- table(df$wind_direction_category)


pie(category_counts, 
    main = "Wind Direction Distribution During the Week of January 8th",
    col = rainbow(length(category_counts)),
    labels = paste0(names(category_counts), ": ", category_counts))


```

# Create a Pie Chart for Wind Direction (Week of Jan 8) Using ggplot2

```{r}

rice |>
   filter(month( Date ) == 1 & day( Date ) >= 7 & day( Date ) <= 13 ) |>
  group_by(Date) |>
  summarize(wind_direction = mean(WindDir, na.rm = TRUE)) -> df
  
breaks <- seq(0, 360, by = 45)  
df$wind_direction_category <- cut(df$wind_direction, breaks = breaks, right = FALSE)

category_counts <- df |>
  group_by(wind_direction_category) |>
  summarize(count = n())

ggplot(category_counts, aes(x = "", y = count, fill = wind_direction_category)) +
  geom_bar(stat = "identity", width = 1) +  
  coord_polar("y") + 
  labs(title = "Wind Direction Distribution During the Week of January 8th",
       fill = "Wind Direction (Degrees)") +  
  theme_minimal()
  scale_fill_brewer(palette = "Set2")  



```
