---
title: "Raster Homework"
author: Kyleigh Johnson
date: 11/5/24
format: html
editor: visual
---

*This is my completion of the raster data homework.*

Rasters are spatial continuous distributions of geospatial data. We have been shown that we an work with data in raster format using either `raster` and working on object directly, or via `dplyr` and using the normal data manipulation routines that we've learned for non-spatial data. In this *in class* exercise, you may use both of these techniques to load, manipulate, and display geospatial data.

## The Raw Data

The data for this will be the same Baja California data we showed in the lectures on [points](https://dyerlab.github.io/ENVS-Lectures/spatial/spatial_points/slides.html#1) and [rasters](https://dyerlab.github.io/ENVS-Lectures/spatial/rasters/slides.html#1).

```{r}
raster_url <- "https://github.com/DyerlabTeaching/Raster-Data/raw/main/data/alt_22.tif"
beetle_url <- "https://raw.githubusercontent.com/DyerlabTeaching/Raster-Data/main/data/AraptusDispersalBias.csv"
```

## The Questions

1.  Load the raster and point data in and crop to an appropriate size to display the locations of the sampling plots and make a label for each site.

```{r}
library( tidyverse )
library( raster )
library( sf ) 

elev <- raster( raster_url )

read_csv( beetle_url )  %>% 
  st_as_sf( coords = c("Longitude","Latitude"), 
            crs = 4326 ) -> beetles 

beetles
```

```{r}
plot( elev, xlab="Longitude" )
```

```{r}
beetles %>%
  st_union() %>%
  st_buffer( dist = 1 ) %>%
  st_bbox() 

baja_extent <- extent( c(-115.29354,-108,22,30))

baja_elev <- crop( elev, baja_extent )
baja_elev
plot( baja_elev) 

```

```{r}
plot( baja_elev, xlab="Longitude", ylab="Latitude")
text( beetles, pch=15, col="red" )
```

```{r}
library( ggrepel )
baja_elev %>%
  rasterToPoints() %>%
  as.data.frame() %>% 
  ggplot() +
  geom_raster( aes(x,y,fill=alt_22) ) +
  geom_sf_label( aes(label=Site), 
                 data = beetles) +
  coord_equal()
```

2.  Use the click() function to crop the raster and filter the sites to include only the sampling locations that are in the Cape regions (e.g., south of site labeled #75). Plot the raster of elevation and all cape region locales with sex-ratio represented as the size of the plot symbol.

```{r}
plot( baja_elev, xlab="Longitude", ylab="Latitude")
text( beetles, pch=16, col="red" )
click( baja_elev, 
       n=2 ) -> points

```

3.  The peninsula of Baja California is divided into the States of *Baja California Norte* and *Baja California Sur*. The border between these states is at 28° Latitude. Divide the sample locations into groups based upon which state they are located and plot the average sex ratio of the sample sites partitioned by each site.

```{r}
beetles <- beetles %>%
  mutate( State = ifelse(st_coordinates(.)[,2] > 28, "Baja California Norte", "Baja California Sur" ) )

average_sex_ratio <- beetles %>%
  group_by( State, Site ) %>%
  summarize( avg_sex_ratio = mean( MFRatio, na.rm = TRUE ), .groups = 'drop' )

baja_elev %>%
  rasterToPoints() %>%
  as.data.frame() %>% 
  ggplot() +
  geom_raster( aes(x = x, y = y, fill = alt_22 ) ) +
  geom_sf(data = beetles, aes( color = State, geometry = geometry ), size = 3) +
  geom_text( data = average_sex_ratio, aes( x = st_coordinates( geometry )[,1],
                                      y = st_coordinates( geometry )[,2],
                                      label = round( avg_sex_ratio, 2)),
             color = "white", size = 3) +
  coord_sf() +
  labs(fill = "Elevation", color = "State" ) +
  theme_minimal()



```

4.  Is there a relationship between the observed sex ratio and the elevation at that site? Plot these, and in the text, perform a correlation test (cor.test()) and report the values as if you were writing them up in a results section.

```{r}
beetles <- beetles %>%
  mutate( Elevation = raster::extract( baja_elev, st_coordinates( beetles ) ) )

correlation_test <- cor.test( beetles$MFRatio, beetles$Elevation )
correlation_test

```

A correlation test was conducted to assess the relationship between sex ratio (MFRatio) and elevation across sample sites. The results indicated a non-significant negative correlation between sex ratio and elevation, t = -0.55949, df = 29, p-value = 0.5801, with a 95% confidence interval \[-0.4415112 0.2605426\], and r = -.103. The negative r value attributes to the negative correlation. Since the p value is well above .05, it is assumed to be non-significant.

This suggests that there is no statistically significant correlation between sex ratio and elevation within the sampled Baja California region.
